---
title: "Hierarchical Calibration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{hierarchical-calibration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(leafblower)
set.seed(42)
```

## Overview

Two-stage hierarchical calibration solves the sparse-cell problem that causes
standard raking to return `NaN` or `RK_ERR_INFEAS` when the joint distribution
of margins is sparse relative to sample size.

**Stage 1** (coarse): calibrate globally on a subset of margins that define
broad cells with adequate counts.

**Stage 2** (fine): within each coarse cell, calibrate on the remaining margins
using only the observations in that cell.

This vignette demonstrates RAKING. Sinkhorn and Greenkhorn examples will be
added in later releases.

---

## Data-generating process: K=9 sparse joint (rescue scenario)

```{r dgp-k9, eval=FALSE}
# K=9 binary margins, N=10 000, Dirichlet(0.1) joint — highly sparse.
# Standard single-stage raking fails on ~80%+ of seeds.
library(leafblower)

make_k9_data <- function(seed, N = 1000L, K = 9L) {
  set.seed(seed)
  # Dirichlet(alpha=0.1) joint probabilities over 2^K cells
  alpha  <- rep(0.1, 2^K)
  probs  <- rgamma(length(alpha), alpha); probs <- probs / sum(probs)
  cells  <- sample(seq_along(probs), N, replace = TRUE, prob = probs)
  # Decode cell index to K binary columns
  df <- as.data.frame(
    matrix(as.integer((outer(cells - 1L, 2^(0:(K-1L)), bitwAnd) > 0L)),
           nrow = N, ncol = K,
           dimnames = list(NULL, paste0("x", seq_len(K)))))
  targets <- lapply(seq_len(K), function(k) c(`0` = 0.5, `1` = 0.5))
  names(targets) <- names(df)
  list(df = df, targets = targets)
}
```

## Single-stage failure demonstration

```{r single-stage-fail, eval=FALSE}
failures <- vapply(1:20, function(s) {
  d   <- make_k9_data(s)
  res <- harvest(d$df, d$targets, method = "raking")
  any(is.nan(res$weight)) || attr(res, "result")$status != 0L
}, logical(1))
cat("Single-stage failure rate:", mean(failures), "\n")
# Expected: >= 0.80
```

## Two-stage rescue demonstration

```{r two-stage-rescue, eval=FALSE}
converged <- vapply(1:20, function(s) {
  d   <- make_k9_data(s)
  res <- harvest(d$df, d$targets, method = "raking",
                 hierarchical = list(
                   coarse_mask      = c(1L, 1L, 1L, 0L, 0L, 0L, 0L, 0L, 0L),
                   min_cell_n       = 30L,
                   mode             = 0L,
                   outer_tol        = 1e-4,
                   outer_iterations = 100L
                 ))
  !any(is.nan(res$weight)) && attr(res, "result")$status == 0L
}, logical(1))
cat("Two-stage convergence rate:", mean(converged), "\n")
# Expected: 1.0
```

## `mode = "exact"` + orthogonality contract

```{r exact-mode, eval=FALSE}
# Exact mode requires that fine-margin levels nest within coarse cells
# (each fine-margin level belongs to exactly one coarse cell).
# Use orthogonal design: coarse = first 2 margins, fine = last 2.
set.seed(1)
N   <- 1000L
df2 <- data.frame(
  region = sample(c("N","S"), N, TRUE),
  sex    = sample(c("M","F"), N, TRUE),
  age    = sample(c("Y","O"), N, TRUE),
  edu    = sample(c("H","L"), N, TRUE)
)
tgt2 <- list(
  region = c(N = 0.5, S = 0.5),
  sex    = c(M = 0.5, F = 0.5),
  age    = c(Y = 0.5, O = 0.5),
  edu    = c(H = 0.5, L = 0.5)
)
# Non-orthogonal split raises BADARG at validation time.
# Orthogonal split (fine levels must be nested within coarse cells) proceeds.
res_exact <- harvest(df2, tgt2, method = "raking",
                     hierarchical = list(
                       coarse_mask      = c(1L, 1L, 0L, 0L),
                       min_cell_n       = 1L,
                       mode             = 1L,
                       outer_tol        = 0,
                       outer_iterations = 1L
                     ))
```

---

*Sinkhorn and Greenkhorn hierarchical examples will be added in T-G and T-H.*
