---
# STUDY-BRANCH-ONLY-DO-NOT-MERGE
title: "leafblower Benchmark Study — Draft Report"
subtitle: "Assembled report (WU-12.13.6) — figures/tables wired; not yet rendered/frozen"
format:
  typst:
    toc: true
bibliography: references.bib
---

```{r}
#| label: setup
#| include: false
source("_data.R")
metrics    <- read_metrics()
agreement  <- read_agreement()
runs       <- read_runs()
```

## Introduction

_Study scope, tiering, and build-variant discipline per `docs/benchmark/DESIGN.md`._

## Work-precision {#sec-work-precision}

_Achieved accuracy vs. wall time, per objective family (`strata`)._ Each panel
plots, per (solver, problem), median wall time against that family's native-
divergence error column, with the empirical Pareto envelope (sort-on-error,
running-min-time) drawn as a black step line. This accuracy-vs-cost view
complements the performance-profile methodology of @sec-profiles in the
spirit of the data-profile framework for benchmarking optimization software
[@morewild2009benchmarking].

```{r}
#| label: gen-work-precision
#| include: false
source("fig_work_precision.R")
```

![Work precision — kl family.](figures/work_precision_kl.pdf){#fig-work-precision-kl}

![Work precision — chi2 family.](figures/work_precision_chi2.pdf){#fig-work-precision-chi2}

![Work precision — logit family.](figures/work_precision_logit.pdf){#fig-work-precision-logit}

![Work precision — minimax family (`margin_linf` used as a NEUTRAL error axis only — Chebyshev/minimax solvers are home-field advantaged on it; see the figure's own caption).](figures/work_precision_minimax.pdf){#fig-work-precision-minimax}

## Wall-time {#sec-wall-time}

_Per-solver median end-to-end wall time, per objective family, with
leafblower's headline (`build_series(..., "headline")`, portable build) and
native ("tuned-build delta") variants distinguished by point shape within
the same panel — not by separate figures._

```{r}
#| label: gen-wall-time
#| include: false
source("fig_walltime.R")
```

![Wall time — kl family.](figures/walltime_kl.pdf){#fig-wall-time-kl}

![Wall time — chi2 family.](figures/walltime_chi2.pdf){#fig-wall-time-chi2}

![Wall time — logit family.](figures/walltime_logit.pdf){#fig-wall-time-logit}

![Wall time — minimax family.](figures/walltime_minimax.pdf){#fig-wall-time-minimax}

## Performance profiles {#sec-profiles}

_Dolan–Moré performance profiles [@dolan2002benchmarking] over the §3C
parametric instance family (`if_*` problems), computed family-stratified per
Gould and Scott's problem-set-sensitivity guidance [@gouldscott2016note],
each with a bootstrap-over-problems pointwise 95% CI ribbon (B ≥ 1000). A
stratum below the |P| < 20 validity floor is flagged in its own caption,
not silently rendered as reliable._

```{r}
#| label: gen-profiles
#| include: false
source("fig_profiles.R")
```

![Performance profile — kl family (headline/portable).](figures/profile_kl.pdf){#fig-profiles-kl}

![Performance profile — chi2 family (headline/portable).](figures/profile_chi2.pdf){#fig-profiles-chi2}

![Performance profile — logit family (headline/portable).](figures/profile_logit.pdf){#fig-profiles-logit}

![Performance profile — minimax family (headline/portable).](figures/profile_minimax.pdf){#fig-profiles-minimax}

![Performance profile — leafblower native-build delta, faceted by family (competitors excluded; there is no native competitor variant).](figures/profile_native_delta.pdf){#fig-profiles-native-delta}

## Convergence {#sec-convergence}

_Per-iteration convergence trajectories for the 8 instrumented solvers
(chebyshev, greenkhorn, greg, newton_kl, oris, oris_soft, raking, sinkhorn),
one faceted figure per problem instance. Each solver's own internal
stopping-criterion residual/gap proxy (`errRp` / `dual_gap` / `marg_defect`)
is named in its facet strip and is NOT comparable across panels with a
different metric name — see each figure's caption._

```{r}
#| label: gen-convergence
#| include: false
source("fig_convergence.R")
```

![Convergence trajectories — if_n10000_K4_medium_moderate_well_loose_s0.](figures/convergence_if_n10000_K4_medium_moderate_well_loose_s0.pdf){#fig-convergence-if_n10000_K4_medium_moderate_well_loose_s0}

![Convergence trajectories — if_n10000_K9_high_extreme_ill_tight_s0.](figures/convergence_if_n10000_K9_high_extreme_ill_tight_s0.pdf){#fig-convergence-if_n10000_K9_high_extreme_ill_tight_s0}

![Convergence trajectories — stepstone_bounded.](figures/convergence_stepstone_bounded.pdf){#fig-convergence-stepstone_bounded}

![Convergence trajectories — stepstone_unbounded.](figures/convergence_stepstone_unbounded.pdf){#fig-convergence-stepstone_unbounded}

## Peak RSS {#sec-rss}

_Memory working-set (`peak_rss_bytes`, baseline-subtracted, max-over-reps
high-water mark), per (solver, problem), headline build only
(`build_series(..., "headline")`)._

```{r}
#| label: gen-rss
#| include: false
source("fig_rss.R")
```

![Peak RSS by solver and problem.](figures/rss.pdf){#fig-rss-by-solver}

```{r}
#| label: tbl-rss-by-solver-problem
#| tbl-cap: "Peak RSS (MiB) by solver and problem, headline build only, baseline-subtracted."
knitr::kable(utils::read.csv("tables/rss_by_solver_problem.csv"))
```

```{r}
#| label: tbl-rss-by-language-package
#| tbl-cap: "Peak RSS (MiB) rollup by (language, package): max over problems of the per-problem high-water mark."
knitr::kable(utils::read.csv("tables/rss_by_language_package.csv"))
```

## Weight agreement {#sec-agreement}

_RQ5: median Pearson/Spearman correlation and relative max|Δw| of raw weight
vectors vs. the independent convex reference, strictly-convex families (kl,
chi2, logit) only — minimax is EXCLUDED by construction (§6 Blocker G: the
L-infinity LP optimum lies on a face, not a unique vertex) and is instead
judged on objective-value agreement (`rel_gap`; see @tbl-minimax-objval)._

```{r}
#| label: gen-agreement
#| include: false
source("fig_agreement.R")
```

![Weight-vector agreement — kl family (strictly-convex).](figures/agreement_kl.pdf){#fig-agreement-kl}

![Weight-vector agreement — chi2 family (strictly-convex).](figures/agreement_chi2.pdf){#fig-agreement-chi2}

![Weight-vector agreement — logit family (strictly-convex).](figures/agreement_logit.pdf){#fig-agreement-logit}

## Tables {#sec-tables}

_RQ4 quality summary (median + interquartile ESS/DEFF, per family/solver/
tier/stale) and the RQ5 minimax objective-value agreement (`rel_gap`,
headline build only)._

```{r}
#| label: gen-quality
#| include: false
source("tbl_quality.R")
```

```{r}
#| label: tbl-quality
#| tbl-cap: "RQ4 quality: median + IQR of ESS and DEFF (Kish and g-weighted), per (family, solver, tier, stale)."
knitr::kable(utils::read.csv("tables/quality.csv"))
```

```{r}
#| label: tbl-minimax-objval
#| tbl-cap: "RQ5 minimax objective-value agreement (rel_gap median + IQR), per solver pair, headline build only."
knitr::kable(utils::read.csv("tables/minimax_objval.csv"))
```

## References

::: {#refs}
:::
